Human capital mediates natural selection in contemporary humans

### 2022-03-19 why-natural-selection.Rmd
Use alternative plot with shapes

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores that predict lower earnings, education and health also predict lower fertility. Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Circles show a significant linear increase (green) or decrease (red) over time (p < 0.05/33)."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
        fatten = 1.2) +
      facet_wrap(~score_name, ncol = 6) +
      my_hline +
      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c("-" = "circle", "+" = "circle", "o" = "square")) +
      theme(
        line        = element_line(size = 0.5, lineend = "round"),
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

To test more directly for natural selection, we regress respondents' relative 
lifetime reproductive success ($RLRS_i$) on their polygenic scores (PGS):

@@

To motivate our economic model, we split the sample by demographic
and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
We examine socio-demographic variables including income and education, and
family structure variables including age at first live birth, presence of a
partner, and lifetime number of sexual partners. Figure
\@ref(fig:plot-income-educ-level) plots selection effects for each polygenic
score, grouping respondents by age of completing full-time education, and by
household income. Effects are larger and more significant for the lowest income
category, and for the lowest education category. These results are robust to
controlling for respondents' age (Appendix section \@ref(sec:age-control)).

@@

We emphasize that our categories are not necessarily exogenous to the polygenic
scores. For example, in the data, and also in our theoretical model, age at first
live birth is a choice variable, which is endogenous to human-capital-related
polygenic scores (Appendix Figure \@ref(fig:plot-age-flb-dv)). Nevertheless,
differences in equation \@ref(eq:regression) across subgroups constrain the set
of possible explanations for natural selection. As we show below, a simple
economic model predicts the differences we find.

@@

Here $a$ captures the strength of preference for children. $u(\cdot)$ captures
the taste for income, and is increasing and concave. We treat $N$ as continuous,
in line with the literature: this can be thought of as the expected number of
children among similar people. The marginal benefit of an extra child is
$\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an increase in human capital on this
marginal benefit is

@@

Thus, a simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
@becker1981treatise. There is also a literature showing negative correlations
between IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income individuals. However, the majority of effect sizes appear unchanged
over a large span of twentieth-century history (Appendix Tables
\@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)), during
which government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact]. Future
work could test this theory more explicitly. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter
$a$ in the model [cf. @jones2008fertility]. This theory would not explain why
selection effects are smaller at higher incomes and education levels. Indeed,
in the model, the effect of $a$ on fertility becomes stronger at higher levels
of human capital. A third alternative is that traits under selection are linked
to externalizing behaviour and risk-seeking. This might be partially captured by
our parameter $\sigma$, which can be interpreted as a measure of risk aversion
over income. However, a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
However, this measure *does* predict the overall number of children, highly
significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
Given that, and the statistical power we get from our sample size, we believe
that the non-result is real: while risk attitude does predict fertility in the
sample, it is not an important channel for natural selection.

@@

Code for this paper is available at 
https://github.com/hughjonesd/negative-selection.

@@

The UK Biobank data contains information on respondents' number of siblings
(including them), i.e. their parents' number of children. Since respondents'
polygenic scores are equal in expectation to the mean scores of their parents,
we can use this to look at selection effects in the parents' generation. We
estimate equation (\ref{eq:regression}) using parents' RLRS as the dependent
variable. The parents' generation has an additional source of ascertainment
bias: sampling parents of respondents overweights parents who have many
children. For instance, parents of three children will have, on average, three
times more children represented in UK Biobank than parents of one child. Parents
of no children will by definition not be represented. To compensate, we reweight
our preferred weightings (Age/Qualification) by the inverse of *number of
siblings*.

@@

```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children. Sibling estimates weighted by 1/number of siblings."}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term     = pretty_names(term),
        term     = reorder_within(term, estimate, dep.var, order_abs(1)),
        dep.var  = fct_inorder(dep.var),
        reg.type = fct_recode(reg.type, 
                               `Resid. for 100 PCs`  = "controlled", 
                               Unresidualized        = "raw"
                             )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p < 0.05/{{n_regs}})) +
        my_vline + 
        facet_wrap(vars(dep.var), scales = "free") +
        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
        my_labs() +
        pval_scale() +
        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
        my_theme() +
        scale_y_reordered() +
        theme(
          axis.text.x  = element_text(size = 7),
          legend.text  = element_text(size = 8),
          legend.title = element_text(size = 9)
        )

>>>>

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores that predict lower earnings, education and health also predict lower fertility. Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "why-natural-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank. Circles are means for 5-year intervals. Lines are 95% confidence intervals. Solid circles show a significant linear increase (green) or decrease (red) over time (p < 0.05/33)."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_linerange(aes(ymin = score_lo, ymax = score_hi), color = "grey60", 
                       size = 3) +
      geom_point(aes(y = score), size = 2) +
      facet_wrap(~score_name, ncol = 6) +
      my_hline +
      scale_color_manual(values = c("-" = "darkred", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c("-" = "circle", "+" = "circle", "o" = "circle open")) +
      theme(
        line        = element_line(size = 0.5, lineend = "round"),
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

To test more directly for natural selection, we regress respondents' relative 
lifetime reproductive success (RLRS) on their polygenic scores (PGS):

@@

We now show the empirical puzzles which motivate our economic model. Each
concerns differences in the strength of natural selection across different
subgroups in the sample. Splitting the sample by demographic and social
variables, we re-estimate \@ref(eq:regression) within each subgroup. We use
socio-economic variables including income and education, and family structure
variables including age at first live birth, presence of a partner, and lifetime
number of sexual partners.

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category. These results are robust
to controlling for respondents' age (Appendix section \@ref(sec:age-control)).

@@

We emphasize that our categories are not necessarily exogenous to polygenic
scores. For example -- both in the data (Appendix Figure
\@ref(fig:plot-age-flb-dv)) and in our model -- age at first live birth is a
choice variable, which is endogenous to an individual's human capital and
related polygenic scores. Nevertheless, observed differences in equation
\@ref(eq:regression) across subgroups constrain the set of possible
theories; a good theory of contemporary natural selection needs to explain
how these results arise. As we show below, a model based on the economic
theory of fertility can do just that.

@@

Here $a$ captures the strength of preference for children. $u(\cdot)$ captures
the taste for income, and is increasing and concave. We treat $N$ as continuous,
in line with the literature: this can be thought of as the expected number of
children among people with a given $a$, $b$ and $W$. The marginal
benefit of an extra child is $\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an
increase in human capital on this marginal benefit is

@@

Thus, a simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
@becker1981treatise. A related literature shows negative correlations
between IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income individuals. However, the majority of effect sizes appear unchanged
over a large span of twentieth-century history (Appendix Table
\@ref(tab:tbl-change-by-period)), during
which government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact]. Future
work could test this theory more explicitly. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter
$a$ in the model [cf. @jones2008fertility]. This theory would not explain why
selection effects are smaller at higher incomes and education levels. Indeed,
in the model, the effect of $a$ on fertility becomes stronger at higher levels
of human capital. A third alternative is that traits under selection are linked
to externalizing behaviour and risk-seeking. This might be partially captured by
our parameter $\sigma$, which can be interpreted as a measure of risk aversion
over income. However, a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
However, this measure *does* predict the overall number of children, highly
significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
Given that, and the statistical power we get from our sample size, we believe
that the non-result is real: while risk attitude does predict fertility in the
sample, it is not an important channel for natural selection.

@@

Code for this paper is available at 
https://github.com/hughjonesd/why-natural-selection.

@@

The UK Biobank data contains information on respondents' number of siblings
(including them), i.e. their parents' number of children. Since respondents'
polygenic scores are equal in expectation to the mean scores of their parents,
we can use this to look at selection effects in the parents' generation. We
estimate equation (\ref{eq:regression}) using parents' RLRS as the dependent
variable.[^parent-RLRS] The parents' generation has an additional source of ascertainment
bias: sampling parents of respondents overweights parents who have many
children. For instance, parents of three children will have, on average, three
times more children represented in UK Biobank than parents of one child. Parents
of no children will by definition not be represented. To compensate, we reweight
our preferred weightings (Age/Qualification) by the inverse of *number of
siblings*.

[^parent-RLRS]: We don't have data on parents' year of birth for most 
respondents. To create parents' RLRS, we divide respondents' number of siblings 
by the average number of siblings of all respondents born in the same year,
weighting the average by respondents' inverse of number of siblings to 
compensate for ascertainment bias.

@@

```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children. Sibling estimates weighted by 1/number of siblings.", fig.height = 6}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term     = pretty_names(term),
        term     = reorder_within(term, estimate, dep.var, order_abs(1)),
        dep.var  = fct_inorder(dep.var),
        dep.var  = fct_recode(dep.var, 
                                `Respondents' RLRS` = "RLRS",
                                `Parents' RLRS` = "RLRS_parents" 
                              ),
        reg.type = fct_recode(reg.type, 
                               `Resid. for 100 PCs`  = "controlled", 
                               Unresidualized        = "raw"
                             )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p < 0.05/{{n_regs}})) +
        my_vline + 
        facet_grid(rows = vars(dep.var), scales = "free") +
        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
        my_labs() +
        pval_scale() +
        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
        my_theme() +
        scale_y_reordered() +
        theme(
          axis.text.x  = element_text(size = 7),
          legend.text  = element_text(size = 8),
          legend.title = element_text(size = 9)
        )
